---
title: "Assignment 1"
author: "Yasamin Nouri Jelyani"
output:
  html_document:
    toc: yes
    toc_depth: 2
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: '2'
bibliography: a2.bib
csl: "biomed-central.csl"
nocite: '@*'
---

# Introduction


KO: DDX knockout
WT: wild type

Women of African American (AA) ancestry are know to have more risk of being 
diagnosed
with ER-/TN(triple negative) breast cancer, as compared to women in European 
American (EA) ancestry. This study looks at a number of differentially expressed 
microRNA (DEmiRs) to identify miRNA patterns that are potential biomarkers 
of this disease. This study found that out of the 102 ER-subtype related DemiRs, majority 
of them were specific to the race, and only 23 were common between the races (1).
Hence, there are subsets of microRNA that are differentially expressed between EA and
AA women. This information help with better targetted treatment of this type of cancer.

Tissue samples:
58 Breast cancer patient (29 AA, 29 EA)
10 healthy women (5 AA, 5 EA)
<br/>
**Question1: What are the control and test conditions of the dataset?** <br/> 
Control condition: Healthy female breast tissue, for both AA and EA decent individuals <br/> 

test condition: Breast cancer breast tissue for both ER+ and ER- cancer types
and for for AA and EA decent patients.
<br/>

**Question2: Why us the dataset of interest to you?** <br/> 
 Since breast cancer has disproportionately effected the individuals from
 the african american decent in the clinic, it is interesting for 
 me to know the reason for this increased vulnerability to breast cancer 
 in this population. This study helps clearify the effect of differentially expressed
 microRNA among the AA and EA population. This information will help guide treatment 
 and prognosis in the clinic. <br/>
 
 
# Find duplicated genes

# Download the Data
First, I downloaded the GEOmetaDB, and got access to the meta data:
```{r eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")
if(!requireNamespace("knitr", quietly=TRUE))
  install.packages("knitr")
if(! requireNamespace("edgeR", quietly=TRUE))
  BiocManager::install("edgeR")
if (!requireNamespace("GEOquery", quietly=TRUE))
  BiocManager::install("GEOquery")
if (!requireNamespace("biomaRt", quietly = TRUE)){
  install.packages("biomaRt")}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!require("stringr", quietly = TRUE))
    install.packages("stringr")
if (!require("dplyr", quietly = TRUE))
    install.packages("dplyr")

library(GEOmetadb)
library(biomaRt)
if( !file.exists("GEOmetadb.sqlite") ) {
    demo_sqlfile <- getSQLiteFile(destdir = getwd(), destfile = "GEOmetadb.sqlite.gz", type = "demo")
} else {
    demo_sqlfile <- "GEOmetadb.sqlite"
    
}


```


Download the data:

```{r message=FALSE}
#if the file exists, open it
fnames <- "GSE156543/GSE156543_Raw_gene_counts_matrix.txt.gz"

if (! file.exists(fnames)){
  gse <- GEOquery::getGEO("GSE117452", GSEMatrix=FALSE)
  sfiles = GEOquery::getGEOSuppFiles("GSE156543")
  fnames = rownames(sfiles)
}
```




## Geo Description of dataset:
```{r message=FALSE}
gse <- GEOquery::getGEO("GSE156543",GSEMatrix=FALSE)
data.frame(head(GEOquery::Meta(gse)))
knitr::kable(data.frame(head(GEOquery::Meta(gse))), format = "html")
```

Platform title:
```{r warning=FALSE}
current_gpl <- names(GEOquery::GPLList(gse))[1]
current_gpl_info <- GEOquery::Meta(GEOquery::getGEO(current_gpl))
```


**Platform title:** `r current_gpl_info$title`<br /> **Submission data:**`r current_gpl_info$submission_date` <br/> **Last Update data:** `r current_gpl_info$last_update_date` <br/> **Organism:** `r current_gpl_info$organism` <br/> **Number of GEO datasets using this technology:** `r length(current_gpl_info$series_id)` <br/> **Number of GEO samples that use this technology:**`r length(current_gpl_info$sample_id)`<br/>


Getting the expression data 
```{r message=FALSE}
DDX39B_exp = read.delim(fnames[1],header=TRUE, check.names = FALSE) 

knitr::kable(DDX39B_exp[1:5,], format = "html")

```

cleaning the data:
I have unique ENSG IDs, so I will not filter
```{r}
dim(DDX39B_exp) 
#58381 rows, 6 col
colnames(DDX39B_exp) 
#(3 KO and 3 WT)
rownames(DDX39B_exp) 
if("ensemble_ID" %in% colnames(DDX39B_exp))
{
  ;
}else{
  DDX39B_exp <- tibble::rownames_to_column(DDX39B_exp, "VALUE")
  colnames(DDX39B_exp)[1]  <- "ensemble_ID"
}

summarized_gene_counts <- sort(table(DDX39B_exp$ensemble_ID), decreasing = TRUE)
knitr::kable(table(DDX39B_exp[, 1])[1:3], format="html")

```


## Gene name structure
Removing genes that do not follow ensembl nomenclature

```{r}
#check that all ensg are unique and all start with ENSG pattern 
areUniquegenes <- length(DDX39B_exp$ensemble_ID) == length(unique(DDX39B_exp$ensemble_ID))
allValidGenes <- length(grep("ENSG.*", DDX39B_exp$ensemble_ID)) == length(DDX39B_exp$ensemble_ID)


```


In edgeR, it is recommended to remove features without at least 1 read per million
in n of the samples, where n is the size of the smallest group of replicates.

```{r}
#removing the versions  
DDX39B_exp <- tidyr::separate(DDX39B_exp, ensemble_ID, into = c("ensemble_ID", "Second"), "[.]")
DDX39B_exp <- DDX39B_exp[, -2]  
  
cpms = edgeR::cpm(DDX39B_exp[,2:7])
rownames(cpms) <- DDX39B_exp$ensembl_id
keep = rowSums(cpms > 1) >= 3 # I have 6 groups, 3 sample replicates and 3 WT replicates
genes_filtered <- DDX39B_exp[keep,]


dim(genes_filtered)
filtered_results <- data.frame(genes_expression = nrow(DDX39B_exp), validGenes = nrow(DDX39B_exp), genes_filtered = nrow(genes_filtered))
rownames(filtered_results)[1] <- "Number of Genes"
filtered_results %>%
  kableExtra::kbl(caption = "Genes count") %>%
  kableExtra::kable_classic(full_width = F, html_font = "Cambria")
```

Data distribution density plot: 

```{r}
data2plot <- log2(edgeR::cpm(genes_filtered[,2:7]))
counts_density <- apply(log2(edgeR::cpm(genes_filtered[,2:7])), 
 2, density)
 #calculate the limits across all the samples
 xlim <- 0; ylim <- 0
 for (i in 1:length(counts_density)) {
 xlim <- range(c(xlim, counts_density[[i]]$x)); 
 ylim <- range(c(ylim, counts_density[[i]]$y))
 }
 cols <- rainbow(length(counts_density))
 ltys <- rep(1, length(counts_density))
 #plot the first density plot to initialize the plot
 plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n", 
 ylab="Smoothing density of log2-CPM", 
 main="", cex.lab = 0.85)
 #plot each line
 for (i in 1:length(counts_density)) 
 lines(counts_density[[i]], col=cols[i], lty=ltys[i])
 #create legend
 legend("topright", colnames(data2plot), 
 col=cols, lty=ltys, cex=0.75, 
 border ="blue", text.col = "green4", 
 merge = TRUE, bg = "gray90")
```

Data distribution boxplot: 
```{r}

data2plot <- log2(edgeR::cpm(genes_filtered[,2:7]))
boxplot(data2plot, xlab = "Samples", ylab = "log2 CPM", 
 las = 2, cex = 0.5, cex.lab = 0.5,
 cex.axis = 0.5, main = "CA125 RNASeq Samples")
#draw the median on each box plot
abline(h = median(apply(data2plot, 2, median)), 
 col = "green", lwd = 0.6, lty = "dashed")
```

after removing outliers

How to remove outliers?
```{r}

```


Normalization by
distribution method

```{r}
limma::plotMA(log2(DDX39B_exp[,c(4,5)]), ylab="M - ratio log expression", 
 main="DDX39B KO + vs WT sample")
```

#NOT WORKING!! 

```{r}
filtered_data_matrix <- as.matrix(genes_filtered[,2:7])
rownames(filtered_data_matrix) <- genes_filtered$ensemble_ID
d <- edgeR::DGEList(counts=filtered_data_matrix)
d <- edgeR::calcNormFactors(d) 
normalized_counts <- edgeR::cpm(d)
limma::plotMDS(normalized_counts, label=c("test1", "test2", "test3", "test4", "test1", "test1"),
 col = c("darkgreen","blue"))
```


```{r}
colnames(normalized_counts)
```


```{r}
library(biomaRt)
```


```{r}

```


To perform mapping, ensembl_gene_id is selected
#DID NOT WORK!!
```{r}
ensembl <- useMart("ensembl")
ensembl <- useDataset("hsapiens_gene_ensembl",mart=ensembl)
conversion_stash <- "DDX3_id_conversion.rds"
if(file.exists(conversion_stash)){
  my_id_conversion <- readRDS(conversion_stash)
} else {
  gene_mapped <- getBM(attributes = c("ensembl_gene_id","hgnc_symbol"), filters = c("ensembl_gene_id"), values=genes_filtered$ensemble_ID, mart = ensembl)
  saveRDS(gene_mapped, conversion_stash)
}
```


```{r}
genes_filtered$ensemble_ID[1:10]
```


```{r}
length(which(rownames(cpms) %in%
gene_mapped$ensembl_gene_id))
```
Total number of genes with measurements
```{r}
nrow(cpms)
```
Number of genes that could not be mapped
```{r}
nrow(cpms) - length(which(rownames(cpms) %in%
gene_mapped$ensembl_gene_id)) 
```
Merging raw data and mapped genes; presenting data in one metobolic syndrome patient and one healthy person
```{r}
cpms_annot <- merge(gene_mapped,cpms,
by.x = 1, by.y = 0, all.y=TRUE)
knitr::kable(cpms_annot[1:6,c(1,2,3,8)],format = "html")
```
* Another method to check ensembl ID missing genes.
```{r}
length(ensembl_id_missing_gene <- cpms_annot$ensembl_gene_id[
which(is.na(cpms_annot$hgnc_symbol))])
```
Presenting the first 10 ensembl IDs that do not match with HUGO
```{r}
knitr::kable(cpms_annot[which(is.na(cpms_annot$hgnc_symbol)),1:3], format = "pipe")[3:12]
```

Old mapping

```{r}
old_mapping <- merge(genes_filtered,data.frame(ensembl_id_missing_gene), by.x = 1, by.y = 1)
knitr::kable(old_mapping[1:10,], type="pipe")
```




# References

